---
title: '3-state Markov model in R'
subtitle: 'with OWSA, TWSA, PSA'
author: "Amina Babar"
output:
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: sentence
---

\newpage


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = T)

```

Change `eval` to `TRUE` if you want to knit this document.

```{r}
rm(list = ls())  # clear memory (removes all the variables from the workspace)
```

# 01 Load packages

```{r}
if (!require('pacman')) install.packages('pacman'); library(pacman) # use this package to conveniently install other packages
# load (install if required) packages from CRAN
p_load("diagram", "dampack", "reshape2") 
# install_github("DARTH-git/darthtools", force = TRUE) Uncomment if there is a newer version
p_load_gh("DARTH-git/darthtools")
```

# 02 Load functions

```{r}
# all functions are in the darthtools package
```

# 03 Input model parameters

```{r}
## General setup
n_cycles        <- 60                            # number of cycles
v_names_cycles  <- paste("cycle", 0:n_cycles)    # cycle names
v_names_states  <- c("DFS", "R", "D")  # state names
n_states        <- length(v_names_states)        # number of health states 

# Discounting factors
d_c             <- 0.04                          # discount rate for costs 
d_e             <- 0.04                          # discount rate for QALYs

# Strategy names
v_names_str     <- c("Standard of Care",         # store the strategy names
                     "Treatment Exp")
                      
n_str           <- length(v_names_str)           # number of strategies

## Transition probabilities
p_DFSD      <- 0.01  # probability of dying when healthy
p_DFSR_SoC  <- 0.05  # probability of becoming sick when healthy, conditional on surviving, under standard of care
p_DFSR_trtExp <- 0.04  # probability of becoming sick when healthy, conditional on surviving, under treatment A
p_RD      <- 0.1   # probability of dying          

## State rewards
# Costs and utilities  
c_DFS       <- 400   # cost of one cycle in healthy state
c_R      <- 1000  # cost of one cycle in sick state
c_D       <- 0     # cost of one cycle in dead state
c_trtExp    <- 800   # cost of treatment A (per cycle) in healthy state
c_SoC       <- 300 
u_DFS       <- 1     # utility when healthy 
u_R       <- 0.5   # utility when sick
u_D       <- 0     # utility when dead
d_e       <- 0.04  # discount rate per cycle equal discount of costs and QALYs by 3%
d_c       <- 0.04  # discount rate per cycle equal discount of costs and QAL  

# Discount weight (equal discounting is assumed for costs and effects)
v_dwc <- 1 / (1 + d_c) ^ (0:n_cycles) 
v_dwe <- 1 / (1 + d_e) ^ (0:n_cycles) 
```

## Draw the state-transition cohort model

```{r}
m_P_diag <- matrix(0, nrow = n_states, ncol = n_states, dimnames = list(v_names_states, v_names_states))
m_P_diag["DFS", "R" ]     = "" 
m_P_diag["DFS", "D" ]     = ""
m_P_diag["DFS", "DFS" ]  = ""
m_P_diag["R"   , "D" ]     = ""
m_P_diag["R"   , "R" ]     = ""
m_P_diag["D"   , "D" ]     = ""
layout.fig <- c(2, 1)
plotmat(t(m_P_diag), t(layout.fig), self.cex = 0.5, curve = 0, arr.pos = 0.8,  
        latex = T, arr.type = "curved", relsize = 0.85, box.prop = 0.8, 
        cex = 0.8, box.cex = 0.7, lwd = 1)
```

# 04 Define and initialize matrices and vectors

## 04.1 Cohort trace

```{r}
## Initial state vector
# All starting healthy
v_s_init <- c("DFS" = 1, "R" = 0, "D" = 0)  
v_s_init

## Initialize cohort trace for cSTM for all strategies
m_M_SoC <- matrix(0, 
                  nrow = (n_cycles + 1), ncol = n_states, 
                  dimnames = list(v_names_cycles, v_names_states))
# Store the initial state vector in the first row of the cohort trace
m_M_SoC[1, ] <- v_s_init
## Initialize cohort traces
m_M_trtExp <- m_M_SoC # structure and initial states remain the same
```

## 04.2 Transition probability matrix

```{r}
## Initialize transition probability matrix 
# all transitions to a non-death state are assumed to be conditional on survival 
m_P_SoC  <- matrix(0,
                   nrow = n_states, ncol = n_states,
                   dimnames = list(v_names_states, v_names_states)) # define row and column names
m_P_SoC
```

Fill in the transition probability matrix:

```{r}
## Standard of Care
# from Healthy
m_P_SoC["DFS", "DFS"] <- (1 - p_DFSD) * (1 - p_DFSR_SoC)
m_P_SoC["DFS", "R"]    <- (1 - p_DFSD) *      p_DFSR_trtExp
m_P_SoC["DFS", "D"]    <-      p_DFSD

# from Sick
m_P_SoC["R", "R"] <- 1 - p_RD
m_P_SoC["R", "D"] <-     p_RD

# from Dead
m_P_SoC["D", "D"] <- 1

## Treatment Exp
m_P_trtExp <- m_P_SoC
m_P_trtExp["DFS", "DFS"] <- (1 - p_DFSD) * (1 - p_DFSR_trtExp)
m_P_trtExp["DFS", "R"]    <- (1 - p_DFSD) *      p_DFSR_trtExp

```

Check if transition probability matrices are valid.

```{r}
# Check that transition probabilities are in [0, 1]
check_transition_probability(m_P_SoC,  verbose = TRUE)
check_transition_probability(m_P_trtExp, verbose = TRUE)

# Check that all rows sum to 1
check_sum_of_transition_array(m_P_SoC,  n_states = n_states, verbose = TRUE)
check_sum_of_transition_array(m_P_trtExp, n_states = n_states, verbose = TRUE)

```

# 05 Run Markov model

```{r}
for (t in 1:n_cycles){  # loop through the number of cycles
  m_M_SoC [t + 1, ] <- m_M_SoC [t, ] %*% m_P_SoC   # estimate the state vector for the next cycle (t + 1)
  m_M_trtExp[t + 1, ] <- m_M_trtExp[t, ] %*% m_P_trtExp  # estimate the state vector for the next cycle (t + 1)
}
head(m_M_SoC)  # print the first few lines of the matrix
```

# 06 Compute and Plot Epidemiological Outcomes

## 06.1 Cohort trace

```{r}
matplot(m_M_SoC, type = 'l', 
        ylab = "Probability of state occupancy",
        xlab = "Cycle",
        main = "Cohort Trace", lwd = 3)  # create a plot of the data
legend("right", v_names_states, col = c("black", "red", "green"), 
       lty = 1:3, bty = "n")  # add a legend to the graph

# plot a vertical line that helps identifying at which cycle the prevalence of sick is highest
abline(v = which.max(m_M_SoC[, "DFS"]), col = "gray")  
```

## 06.2 Overall Survival (OS)

```{r}
v_os <- 1 - m_M_SoC[, "D"]    # calculate the overall survival (OS) probability
v_os <- rowSums(m_M_SoC[, 1:2])  # alternative way of calculating the OS probability   

plot(v_os, type = 'l', 
     ylim = c(0, 1),
     ylab = "Survival probability",
     xlab = "Cycle",
     main = "Overall Survival")  # create a simple plot showing the OS

# add grid 
grid(nx = n_cycles, ny = 10, col = "lightgray", lty = "dotted", lwd = par("lwd"), 
     equilogs = TRUE) 
```

## 06.2.1 Life Expectancy (LE)

```{r}
v_le <- sum(v_os)  # summing probability of OS over time  (i.e. life expectancy)
```

## 06.3 Disease prevalence

```{r}
v_prev <- m_M_SoC[, "R"]/v_os
plot(v_prev,
     ylim = c(0, 1),
     ylab = "Prevalence",
     xlab = "Cycle",
     main = "Disease prevalence")
```

# 07 Compute Cost-Effectiveness Outcomes

## 07.1 Mean Costs and QALYs

```{r}
# per cycle
# calculate expected costs by multiplying cohort trace with the cost vector for the different health states   
v_tc_SoC  <- m_M_SoC  %*% c(c_DFS + c_SoC, c_R, c_D)  
v_tc_trtExp <- m_M_trtExp %*% c(c_DFS + c_trtExp, c_R, c_D)  

# calculate expected QALYs by multiplying cohort trace with the utilities for the different health states   
v_tu_SoC  <- m_M_SoC  %*% c(u_DFS, u_R, u_D)  
v_tu_trtExp <- m_M_trtExp %*% c(u_DFS, u_R, u_D) 

```

## 07.2 Discounted Mean Costs and QALYs

```{r}
# Discount costs by multiplying the cost vector with discount weights (v_dwc) 
tc_d_SoC  <-  t(v_tc_SoC)  %*% v_dwc
tc_d_trtExp <-  t(v_tc_trtExp) %*% v_dwc

# Discount QALYS by multiplying the QALYs vector with discount weights (v_dwe)
tu_d_SoC  <-  t(v_tu_SoC)  %*% v_dwe
tu_d_trtExp <-  t(v_tu_trtExp) %*% v_dwe


# Store them into a vector
v_tc_d <- c(tc_d_SoC, tc_d_trtExp)
v_tu_d <- c(tu_d_SoC, tu_d_trtExp)

# Dataframe with discounted costs and effectiveness
df_ce <- data.frame(Strategy = v_names_str,
                    Cost     = v_tc_d, 
                    Effect   = v_tu_d)
df_ce
```

## 07.3 Compute ICERs of the Markov model

```{r}
df_cea <- calculate_icers(cost       = df_ce$Cost,
                          effect     = df_ce$Effect,
                          strategies = df_ce$Strategy
                          )
df_cea
```

## 07.4 Plot frontier of the Markov model

```{r}
plot(df_cea, effect_units = "QALYs")
```

# 08 Deterministic Sensitivity Analysis

## 08.1 List of input parameters

Create list `l_params_all` with all input probabilities, cost and utilities.

```{r}
l_params_all <- as.list(data.frame(
  p_DFSD      = p_DFSD,  # probability of dying when healthy
  p_DFSR_SoC  = p_DFSR_SoC,  # probability of becoming sick when healthy, under standard of care
  p_DFSR_trtExp = p_DFSR_trtExp,  # probability of becoming sick when healthy, under treatment A
  p_RD      = p_RD,   # probability of dying when sick
  c_DFS       = c_DFS,   # cost of one cycle in healthy state
  c_SoC   = c_SoC,
  c_R       = c_R,  # cost of one cycle in sick state
  c_D       = c_D,     # cost of one cycle in dead state
  c_trtExp    = c_trtExp,   # cost of treatment A (per cycle)
  u_DFS       = u_DFS,     # utility when healthy 
  u_R       = u_R,   # utility when sick
  u_D       = u_D,     # utility when dead
  d_e       = d_e,  # discount factor for effectiveness
  d_c       = d_c   # discount factor for costs
))

# store the parameter names into a vector
v_names_params <- names(l_params_all)
```

## 08.2 Load Sick-Sicker Markov model function

```{r}
source("Functions_markov_3state.R")
# Test function
calculate_ce_out(l_params_all)
```

## 08.3 One-way sensitivity analysis (OWSA)

```{r}
options(scipen = 999) # disabling scientific notation in R
# dataframe containing all parameters, their base case values, and the min and 
# max values of the parameters of interest 
df_params_owsa <- data.frame(pars = c("c_trtExp", "c_SoC", "c_R", "c_DFS", "u_DFS", "u_R", "p_DFSR_SoC", "p_DFSR_trtExp", "p_DFSD", "p_RD"),
                             min  = c(200 ,100, 500, 100, 0.5, 0.1, 0.01, 0.01, 0.001, 0.05 ),  # min parameter values
                             max  = c(1200, 1000, 2000, 1000, 1, 0.9, 0.2, 0.2, 0.2, 0.5)  # max parameter values
                             )

owsa_nmb  <- run_owsa_det(params_range     = df_params_owsa,    # dataframe with parameters for OWSA
                          params_basecase  = l_params_all,      # list with all parameters
                          nsamp            = 100,               # number of parameter values
                          FUN              = calculate_ce_out,  # function to compute outputs
                          outcomes         = c("NMB"),          # output to do the OWSA on
                          strategies       = v_names_str,       # names of the strategies
                          n_wtp            = 5000)              # extra argument to pass to FUN
```

## 08.3.1 Plot OWSA

```{r}
plot(owsa_nmb, txtsize = 10, n_x_ticks = 4, 
     facet_scales = "free") +
     theme(legend.position = "bottom")
```

## 08.3.2 Optimal strategy with OWSA

```{r}
owsa_opt_strat(owsa = owsa_nmb, txtsize = 10)
```

## 08.3.3 Tornado plot

```{r}
owsa_tornado(owsa = owsa_nmb, txtsize = 11)
```

## 08.4 Two-way sensitivity analysis (TWSA)

```{r}
# dataframe containing all parameters, their basecase values, and the min and 
# max values of the parameters of interest
df_params_twsa <- data.frame(pars = c("c_SoC", "c_trtExp"),
                             min  = c(100, 500),  # min parameter values
                             max  = c(6000, 6000) # max parameter values
                             )

twsa_nmb <- run_twsa_det(params_range    = df_params_twsa,    # dataframe with parameters for TWSA
                         params_basecase = l_params_all,      # list with all parameters
                         nsamp           = 40,                # number of parameter values
                         FUN             = calculate_ce_out,  # function to compute outputs
                         outcomes        = "NMB",             # output to do the TWSA on
                         strategies      = v_names_str,       # names of the strategies
                         n_wtp           = 5000)              # extra argument to pass to FUN
```

## 08.4.1 Plot TWSA

```{r}
plot(twsa_nmb)
```

# 09 Probabilistic Sensitivity Analysis (PSA)

```{r}
# Function to generate PSA input dataset
gen_psa <- function(n_sim = 1000, seed = 071818){
  set.seed(seed) # set a seed to be able to reproduce the same results
  df_psa <- data.frame(
    # Transition probabilities (per cycle), conditional on surviving
    # probability to become sick when healthy
     # probability of dying when healthy
    p_DFSD       = rbeta(n_sim, shape1 = 4,  shape2 = 391),
    p_DFSR_SoC   = rbeta(n_sim, shape1 = 24, shape2 = 450),  # under standard of care
    p_DFSR_trtExp  = rbeta(n_sim, shape1 = 15, shape2 = 368),  # under treatment Exp
  

    # probability of dying when sick
    p_DFSD       = rbeta(n_sim, shape1 = 22.4, shape2 = 201.6), 

    # Cost vectors with length n_sim
    # cost of remaining one cycle in state H
    c_DFS        = rgamma(n_sim, shape = 16, scale = 25), 
    # cost of remaining one cycle in state S1
    c_R        = rgamma(n_sim, shape = 100, scale = 10), 
    # cost of being in the death state
    c_D        = 0, 
    # cost of treatment (per cycle)
    c_SoC    = rgamma(n_sim, shape = 64, scale = 12.5),
    # cost of treatment (per cycle)
    c_trtExp    = rgamma(n_sim, shape = 225, scale = 6.67),
    
    # Utility vectors with length n_sim 
    # utility when healthy
    u_DFS        = rbeta(n_sim, shape1 =  1.5, shape2 = 0.0015), 
    # utility when sick
    u_R        = rbeta(n_sim, shape1 = 49.5, shape2 = 49.5), 
    # utility when dead
    u_D        = 0                                              
  )
  return(df_psa)
}


# Try it
gen_psa(10) 

# Number of simulations
n_sim <- 1000

# Generate PSA input dataset
df_psa_input <- gen_psa(n_sim = n_sim)
# First six observations
head(df_psa_input)

# Save the dataframe
# save dataframe
#save(df_psa_input, file = "df_psa_input.rda")


# Histogram of parameters
ggplot(melt(df_psa_input, variable.name = "Parameter"), aes(x = value)) +
       facet_wrap(~Parameter, scales = "free") +
       geom_histogram(aes(y = ..density..)) +
       theme_bw(base_size = 16) + 
       theme(axis.text = element_text(size=8))

# Initialize dataframes with PSA output 
# Dataframe of costs
df_c <- as.data.frame(matrix(0, 
                      nrow = n_sim,
                      ncol = n_str))
colnames(df_c) <- v_names_str
# Dataframe of effectiveness
df_e <- as.data.frame(matrix(0, 
                      nrow = n_sim,
                      ncol = n_str))
colnames(df_e) <- v_names_str
```

## 09.1 Conduct probabilistic sensitivity analysis

```{r}
# Run Markov model on each parameter set of PSA input dataset
for(i in 1:n_sim){
  l_out_temp <- calculate_ce_out(df_psa_input[i, ])
  df_c[i, ] <- l_out_temp$Cost
  df_e[i, ] <- l_out_temp$Effect
  # Display simulation progress
  if(i/(n_sim/10) == round(i/(n_sim/10), 0)) { # display progress every 10%
    cat('\r', paste(i/n_sim * 100, "% done", sep = " "))
  }
}
```

## 09.2 Create PSA object for dampack

```{r}
l_psa <- make_psa_obj(cost          = df_c, 
                      effectiveness = df_e, 
                      parameters    = df_psa_input, 
                      strategies    = v_names_str)
```

## 09.2.1 Save PSA objects

```{r}
save(df_psa_input, df_c, df_e, v_names_str, n_str, l_psa,
     file = "markov_3state_PSA_dataset.RData")
```

Vector with willingness-to-pay (WTP) thresholds.

```{r}
v_wtp <- seq(0, 30000, by = 1000)
```

## 09.3.1 Cost-Effectiveness Scatter plot

```{r}
plot(l_psa, xlim = c(9.5, 22.5))
```

## 09.4 Conduct CEA with probabilistic output

```{r}
# Compute expected costs and effects for each strategy from the PSA
df_out_ce_psa <- summary(l_psa)

# Calculate incremental cost-effectiveness ratios (ICERs)
df_cea_psa <- calculate_icers(cost       = df_out_ce_psa$meanCost, 
                              effect     = df_out_ce_psa$meanEffect,
                              strategies = df_out_ce_psa$Strategy)
df_cea_psa

# Save CEA table with ICERs
# As .RData
save(df_cea_psa, 
     file = "markov_3state_probabilistic_CEA_results.RData")
# As .csv
write.csv(df_cea_psa, 
          file = "markov_3state_probabilistic_CEA_results.csv")
```

## 09.4.1 Plot cost-effectiveness frontier

```{r}
plot(df_cea_psa)
```

## 09.4.2 Cost-effectiveness acceptability curves (CEACs) and frontier (CEAF)

```{r}
ceac_obj <- ceac(wtp = v_wtp, psa = l_psa)
# Regions of highest probability of cost-effectiveness for each strategy
summary(ceac_obj)
# CEAC & CEAF plot
plot(ceac_obj)
```

## 09.4.3 Expected Loss Curves (ELCs)

The expected loss is the the quantification of the foregone benefits when choosing a suboptimal strategy given current evidence.

```{r}
elc_obj <- calc_exp_loss(wtp = v_wtp, psa = l_psa)
elc_obj
# ELC plot
plot(elc_obj, log_y = FALSE)
```

## 09.4.4 Expected value of perfect information (EVPI)

```{r}
evpi <- calc_evpi(wtp = v_wtp, psa = l_psa)
# EVPI plot
plot(evpi, effect_units = "QALY")
```
