---
title: '3-state Markov model in R'
subtitle: 'with OWSA, TWSA, PSA'
author: "Amina Babar"
output:
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: sentence
---
```{r}
rm(list = ls())  # clear memory (removes all the variables from the workspace)
```

# 01 Load packages

```{r}
if (!require('pacman')) install.packages('pacman'); library(pacman) # use this package to conveniently install other packages
# load (install if required) packages from CRAN
p_load("diagram", "dampack", "reshape2") 
# install_github("DARTH-git/darthtools", force = TRUE) Uncomment if there is a newer version
p_load_gh("DARTH-git/darthtools")
```

# 02 Load functions

```{r}
# all functions are in the darthtools package
```

Input model parameters

```{r}
## General setup
n_cycles        <- 35 # number of cycles
v_names_cycles  <- paste("cycle", 0:n_cycles) # cycle names
v_names_states  <- c("PFS", "OS", "D")  # state names
n_states        <- length(v_names_states) # number of health states 

# Discounting factors
d_c             <- 0.04 # discount rate for costs 
d_e             <- 0.04 # discount rate for QALYs

# Strategy names
v_names_str     <- c("Standard of Care",         # store the strategy names
                     "Treatment Exp")
                      
n_str           <- length(v_names_str)           # number of strategies

# transform rate to probability

# probability to move from one state to the next under experimental treatment 
r_PFSD_trtExp <-0.39
t <- 1
p_PFSD_trtExp <-1 - exp(-r_PFSD_trtExp*t)

r_PFSOS_trtExp <-0.333
t <- 1
p_PFSOS_trtExp <-1 - exp(-r_PFSOS_trtExp*t)
p_PFSOS_trtExp



r_OSD_SoC <-0.401
t <- 1
p_OSD_SoC <-1 - exp(-r_OSD_SoC*t)

hr_PFSOS <- 1.22
#1.22
hr_PFSD <- 1.25

# probability to move from one state to the next under SoC 
p_PFSOS_SoC<-(p_PFSOS_trtExp)*(hr_PFSOS) 
p_PFSOS_SoC
rnew<- log(1-(p_PFSOS_SoC))
rnew
#r_PFSOS_soc <- ln(p_PFSD_SoC)
p_PFSD_SoC<-(p_PFSD_trtExp)*(hr_PFSD)
p_OSD_trtExp<-p_OSD_SoC

#State rewards
# Costs and utilities  
c_PFS       <- 723.51  # cost of one cycle in PFS (admin costs)
c_OS      <- 6543.51 # cost of one cycle in OS (admin costs)
c_D       <- 0     # cost of one cycle in dead state
c_trtExp   <- 5998   # direct cost of experimental treatment 
c_SoC       <- 6068.7 #direct cost of SoC
u_PFS       <- 0.7     # utility of PFS per cycle  
u_OS       <- 0.5   # utility of OS per cycle
u_D       <- 0     # utility when dead
d_e       <- 0.04  # discount rate per cycle equal discount of costs and QALYs 
d_c       <- 0.04  # discount rate per cycle equal discount of costs and QALYs  

# Discount weight (equal discounting is assumed for costs and effects)
v_dwc <- 1 / (1 + d_c) ^ (0:n_cycles) 
v_dwe <- 1 / (1 + d_e) ^ (0:n_cycles) 
```

Draw the state-transition cohort model

```{r}
m_P_diag <- matrix(0, nrow = n_states, ncol = n_states, dimnames = list(v_names_states, v_names_states))
m_P_diag["PFS", "OS" ]     = "" 
m_P_diag["PFS", "D" ]     = ""
m_P_diag["PFS", "PFS" ]  = ""
m_P_diag["OS"   , "D" ]     = ""
m_P_diag["OS"   , "OS" ]     = ""
m_P_diag["D"   , "D" ]     = ""
layout.fig <- c(2, 1)
plotmat(t(m_P_diag), t(layout.fig), self.cex = 0.5, curve = 0, arr.pos = 0.8,  
        latex = T, arr.type = "curved", relsize = 0.85, box.prop = 0.8, 
        cex = 0.8, box.cex = 0.7, lwd = 1)
```

Define and initialize matrices and vectors
Cohort trace

```{r}
#Initial state vector
# All starting in PFS
v_s_init <- c("PFS" = 1, "OS" = 0, "D" = 0)  
v_s_init

## Initialize cohort trace for cSTM for all strategies
m_M_SoC <- matrix(0, 
                  nrow = (n_cycles + 1), ncol = n_states, 
                  dimnames = list(v_names_cycles, v_names_states))
# Store the initial state vector in the first row of the cohort trace
m_M_SoC[1, ] <- v_s_init
## Initialize cohort traces
m_M_trtExp <- m_M_SoC # structure and initial states remain the same
```

Transition probability matrix

```{r}
# Initialize transition probability matrix 
# all transitions to a non-death state are assumed to be conditional on survival 
m_P_SoC  <- matrix(0,
                   nrow = n_states, ncol = n_states,
                   dimnames = list(v_names_states, v_names_states)) # define row and column names
m_P_SoC
```

Fill in the transition probability matrix

```{r}
## Standard of Care
# from PFS
m_P_SoC["PFS", "PFS"] <- (1 - p_PFSD_SoC) * (1 - p_PFSOS_SoC)
m_P_SoC["PFS", "OS"]    <- (1 - p_PFSD_SoC) * (p_PFSOS_SoC)
m_P_SoC["PFS", "D"]    <- p_PFSD_SoC

# from OS
m_P_SoC["OS", "OS"] <- 1 - p_OSD_SoC
m_P_SoC["OS", "D"] <- p_OSD_SoC

# from D
m_P_SoC["D", "D"] <- 1

m_P_trtExp <- m_P_SoC

## Treatment Exp

m_P_trtExp["PFS", "PFS"] <- (1 - p_PFSD_trtExp) * (1 - p_PFSOS_trtExp)
m_P_trtExp["PFS", "OS"]    <- (1 - p_PFSD_trtExp) * (p_PFSOS_trtExp)
m_P_trtExp["PFS", "D"] <- p_PFSD_trtExp
m_P_trtExp["D", "D"] <- 1

```

For a transition matrix to be valid, each row must be a probability vector, and the sum of all its terms must be 1.To check if transition probability matrices are valid we run the following command to confirm their validity by checking if each row equals to 1. 

```{r}
# Check that transition probabilities are in [0, 1]
check_transition_probability(m_P_SoC,  verbose = TRUE)
check_transition_probability(m_P_trtExp, verbose = TRUE)

# Check that all rows sum to 1
check_sum_of_transition_array(m_P_SoC,  n_states = n_states, verbose = TRUE)
check_sum_of_transition_array(m_P_trtExp, n_states = n_states, verbose = TRUE)

```

Markov model

```{r}
for (t in 1:n_cycles){  # loop through the number of cycles
  m_M_SoC [t + 1, ] <- m_M_SoC [t, ] %*% m_P_SoC   # estimate the state vector for the next cycle (t + 1)
  m_M_trtExp[t + 1, ] <- m_M_trtExp[t, ] %*% m_P_trtExp  # estimate the state vector for the next cycle (t + 1)
}
head(m_M_SoC)  # print the first few lines of the matrix
```

Epidemiological Outcomes
Cohort trace

```{r}
matplot(m_M_SoC, type = 'l', 
        ylab = "Probability of state occupancy",
        xlab = "Cycle",
        main = "Cohort Trace", lwd = 3)  # create a plot of the data
legend("right", v_names_states, col = c("black", "red", "green"), 
       lty = 1:3, bty = "n")  # add a legend to the graph

# plot a vertical line that helps identifying at which cycle the prevalence of sick is highest
abline(v = which.max(m_M_SoC[, "PFS"]), col = "gray")  
```

Overall Survival (OS)

```{r}
v_os <- 1 - m_M_SoC[, "D"]    # calculate the overall survival (OS) probability
v_os <- rowSums(m_M_SoC[, 1:2])  # alternative way of calculating the OS probability   

plot(v_os, type = 'l', 
     ylim = c(0, 1),
     ylab = "Survival probability",
     xlab = "Cycle",
     main = "Overall Survival")  # create a simple plot showing the OS

# add grid 
grid(nx = n_cycles, ny = 10, col = "lightgray", lty = "dotted", lwd = par("lwd"), 
     equilogs = TRUE) 
```

Life Expectancy (LE)

```{r}
v_le <- sum(v_os)  # summing probability of OS over time  (i.e. life expectancy)
```

Disease prevalence

```{r}
v_prev <- m_M_SoC[, "OS"]/v_os
plot(v_prev,
     ylim = c(0, 1),
     ylab = "Prevalence",
     xlab = "Cycle",
     main = "Disease prevalence")
```

Cost-Effectiveness Outcomes

Mean Costs and QALYs

```{r}
# per cycle
# calculate expected costs by multiplying cohort trace with the cost vector for the different health states   
v_tc_SoC  <- m_M_SoC  %*% c(c_PFS + c_SoC, c_OS, c_D)  
v_tc_trtExp <- m_M_trtExp %*% c(c_PFS + c_trtExp, c_OS, c_D)  

# calculate expected QALYs by multiplying cohort trace with the utilities for the different health states   
v_tu_SoC  <- m_M_SoC  %*% c(u_PFS, u_OS, u_D)  
v_tu_trtExp <- m_M_trtExp %*% c(u_PFS, u_OS, u_D) 

```
Discounted Mean Costs and QALYs

In order to determine the net present value of future expenditures and benefits, we used a discount rate of 4%. Using this rate, we next create a column vector with cycle-specific discount weights d of size nT+1, where the tth element reflects the discounting for cycle t. The following formula and code corresponds to the discounting of SoC and experimental treatment total costs and utilities (QALYS). Table 5 shows the discounted values of these costs and effects.

d =1,1/〖(1+d)〗^1 ,1/〖(1+d)〗^2   ,··· ,1/〖(1+d)〗^nT   


```{r}
# Discount costs by multiplying the cost vector with discount weights (v_dwc) 
tc_d_SoC  <-  t(v_tc_SoC)  %*% v_dwc
tc_d_trtExp <-  t(v_tc_trtExp) %*% v_dwc

# Discount QALYS by multiplying the QALYs vector with discount weights (v_dwe)
tu_d_SoC  <-  t(v_tu_SoC)  %*% v_dwe
tu_d_trtExp <-  t(v_tu_trtExp) %*% v_dwe


# Store them into a vector
v_tc_d <- c(tc_d_SoC, tc_d_trtExp)
v_tu_d <- c(tu_d_SoC, tu_d_trtExp)

# Dataframe with discounted costs and effectiveness
df_ce <- data.frame(Strategy = v_names_str,
                    Cost     = v_tc_d, 
                    Effect   = v_tu_d)
df_ce
```

Compute ICERs of the Markov model

```{r}
df_cea <- calculate_icers(cost       = df_ce$Cost,
                          effect     = df_ce$Effect,
                          strategies = df_ce$Strategy
                          )
df_cea
```

Plot frontier of the Markov model

```{r}
plot(df_cea, effect_units = "QALYs")
```

# 08 Deterministic Sensitivity Analysis

## 08.1 List of input parameters

Create list `l_params_all` with all input probabilities, cost and utilities.

```{r}
l_params_all <- as.list(data.frame(
  p_PFSOS_SoC  = p_PFSOS_SoC,  
  p_PFSOS_trtExp = p_PFSOS_trtExp,  
  p_OSD_trtExp      = p_OSD_trtExp,   
  p_PFSD_trtExp= p_PFSD_trtExp,
  c_PFS       = c_PFS,   
  c_SoC   = c_SoC,
  c_OS       = c_OS,  
  c_D       = c_D,     
  c_trtExp    = c_trtExp,   
  c_SoC       = c_SoC, 
  u_PFS       = u_PFS,      
  u_OS       = u_OS,   
  u_D       = u_D,     
  d_e       = d_e,  
  d_c       = d_c   
))

# store the parameter names into a vector
v_names_params <- names(l_params_all)
```

## 08.2 Load  Markov model function

```{r}
source("Functions_markov_3state.R")
# Test function
calculate_ce_out(l_params_all)
```

## 08.3 One-way sensitivity analysis (OWSA)

```{r}
options(scipen = 999) # disabling scientific notation in R
# dataframe containing all parameters, their base case values, and the min and 
# max values of the parameters of interest 
df_params_owsa <- data.frame(pars = c("c_trtExp", "c_SoC", "c_OS", "c_PFS", "u_PFS", "u_OS", "p_PFSOS_SoC", "p_PFSOS_trtExp", "p_PFSD_trtExp", "p_OSD_trtExp"),
                             min  = c(4800 ,4800, 5235, 580, 0.56, 0.4, 0.28, 0.224, 0.258, 0.264 ),  # min parameter values
                             max  = c(7200, 7200, 7853, 875, 0.84, 0.6, 0.42, 0.336, 0.388, 0.396)  # max parameter values
                             )

owsa_nmb  <- run_owsa_det(params_range     = df_params_owsa,    # dataframe with parameters for OWSA
                          params_basecase  = l_params_all,      # list with all parameters
                          nsamp            = 100,               # number of parameter values
                          FUN              = calculate_ce_out,  # function to compute outputs
                          outcomes         = c("NMB"),          # output to do the OWSA on
                          strategies       = v_names_str,       # names of the strategies
                          n_wtp            = 20000)              # extra argument to pass to FUN
```

## 08.3.1 Plot OWSA

```{r}
plot(owsa_nmb, txtsize = 8, n_x_ticks =4, 
     facet_scales = "free") +
     theme(legend.position = "bottom")
```


## 08.3.3 Tornado plot

```{r}
owsa_tornado(owsa = owsa_nmb, txtsize = 11)
```

## 08.4 Two-way sensitivity analysis (TWSA)

```{r}
# dataframe containing all parameters, their basecase values, and the min and 
# max values of the parameters of interest
df_params_twsa <- data.frame(pars = c("c_SoC", "c_trtExp"),
                             min  = c(1000, 1000),  # min parameter values
                             max  = c(10000, 10000) # max parameter values
                             )

twsa_nmb <- run_twsa_det(params_range    = df_params_twsa,    # dataframe with parameters for TWSA
                         params_basecase = l_params_all,      # list with all parameters
                         nsamp           = 40,                # number of parameter values
                         FUN             = calculate_ce_out,  # function to compute outputs
                         outcomes        = "NMB",             # output to do the TWSA on
                         strategies      = v_names_str,       # names of the strategies
                         n_wtp           = 20000)              # extra argument to pass to FUN
```

## 08.4.1 Plot TWSA

```{r}
plot(twsa_nmb)
```

# 09 Probabilistic Sensitivity Analysis (PSA)

```{r}
# Function to generate PSA input dataset
gen_psa <- function(n_sim = 10000, seed = 123){
  set.seed(seed) # set a seed to be able to reproduce the same results
  df_psa <- data.frame(
    # Transition probabilities (per cycle), conditional on surviving
    # probability to become sick when healthy
     # probability of dying when healthy
    p_PFSD       = rbeta(n_sim, shape1 = 11,  shape2 = 36), # 566 participants (death rate=0.231)
    p_PFSOS_SoC   = rbeta(n_sim, shape1 = 20, shape2 = 27),  # 566 participants (progression=0.437), new=0.423- stay:240 leave:326
    p_PFSOS_trtExp  = rbeta(n_sim, shape1 = 23, shape2 = 24),  # 566 participants (progression=0.499)

    # probability of dying when sick
    p_PFSD       = rbeta(n_sim, shape1 = 121, shape2 = 415), 

    # Cost vectors with length n_sim
    # cost of remaining one cycle in state H
    c_PFS        = rgamma(n_sim, shape = 100, scale = 65.4351), 
    # cost of remaining one cycle in state S1
    c_OS        = rgamma(n_sim, shape = 100, scale = 7.24), 
    # cost of being in the death state
    c_D        = 0, 
    # cost of treatment (per cycle)
    c_SoC    = rgamma(n_sim, shape = 100, scale = 60.69),
    # cost of treatment (per cycle)
    c_trtExp    = rgamma(n_sim, shape = 100, scale = 59.98 ),
    
    # Utility vectors with length n_sim 
    # utility when healthy
    u_PFS        = rbeta(n_sim, shape1 =  100, shape2 = 0.007), 
    # utility when sick
    u_OS        = rbeta(n_sim, shape1 = 100, shape2 = 0.005), 
    # utility when dead
    u_D        = 0                                              
  )
  return(df_psa)
}


# Try it
gen_psa(10) 

# Number of simulations
n_sim <- 10000

# Generate PSA input dataset
df_psa_input <- gen_psa(n_sim = n_sim)
# First six observations
head(df_psa_input)

# Save the dataframe
# save dataframe
#save(df_psa_input, file = "df_psa_input.rda")


# Histogram of parameters
ggplot(melt(df_psa_input, variable.name = "Parameter"), aes(x = value)) +
       facet_wrap(~Parameter, scales = "free") +
       geom_histogram(aes(y = ..density..)) +
       theme_bw(base_size = 12) + 
       theme(axis.text = element_text(size=6))

# Initialize dataframes with PSA output 
# Dataframe of costs
df_c <- as.data.frame(matrix(0, 
                      nrow = n_sim,
                      ncol = n_str))
colnames(df_c) <- v_names_str
# Dataframe of effectiveness
df_e <- as.data.frame(matrix(0, 
                      nrow = n_sim,
                      ncol = n_str))
colnames(df_e) <- v_names_str
```

## 09.1 Conduct probabilistic sensitivity analysis

```{r}
# Run Markov model on each parameter set of PSA input dataset
for(i in 1:n_sim){
  l_out_temp <- calculate_ce_out(df_psa_input[i, ])
  df_c[i, ] <- l_out_temp$Cost
  df_e[i, ] <- l_out_temp$Effect
  # Display simulation progress
  if(i/(n_sim/10) == round(i/(n_sim/10), 0)) { # display progress every 10%
    cat('\r', paste(i/n_sim * 100, "% done", sep = " "))
  }
}
```

## 09.2 Create PSA object for dampack

```{r}
l_psa <- make_psa_obj(cost          = df_c, 
                      effectiveness = df_e, 
                      parameters    = df_psa_input, 
                      strategies    = v_names_str)
```

## 09.2.1 Save PSA objects

```{r}
save(df_psa_input, df_c, df_e, v_names_str, n_str, l_psa,
     file = "markov_3state_PSA_dataset.RData")
```

Vector with willingness-to-pay (WTP) thresholds.

```{r}
v_wtp <- seq(0, 50000, by = 1000)
```

## 09.3.1 Cost-Effectiveness Scatter plot

```{r}
plot(l_psa, xlim = c(2.4, 3.00))
```

## 09.4 Conduct CEA with probabilistic output

```{r}
# Compute expected costs and effects for each strategy from the PSA
df_out_ce_psa <- summary(l_psa)

# Calculate incremental cost-effectiveness ratios (ICERs)
df_cea_psa <- calculate_icers(cost       = df_out_ce_psa$meanCost, 
                              effect     = df_out_ce_psa$meanEffect,
                              strategies = df_out_ce_psa$Strategy)
df_cea_psa

# Save CEA table with ICERs
# As .RData
save(df_cea_psa, 
     file = "markov_3state_probabilistic_CEA_results.RData")
# As .csv
write.csv(df_cea_psa, 
          file = "markov_3state_probabilistic_CEA_results.csv")
```

## 09.4.1 Plot cost-effectiveness frontier

```{r}
plot(df_cea_psa)
```

## 09.4.2 Cost-effectiveness acceptability curves (CEACs) and frontier (CEAF)

```{r}
ceac_obj <- ceac(wtp = v_wtp, psa = l_psa)
# Regions of highest probability of cost-effectiveness for each strategy
summary(ceac_obj)
# CEAC & CEAF plot
plot(ceac_obj)
```

## 09.4.3 Expected Loss Curves (ELCs)

The expected loss is the the quantification of the foregone benefits when choosing a suboptimal strategy given current evidence.

```{r}
elc_obj <- calc_exp_loss(wtp = v_wtp, psa = l_psa)
elc_obj
# ELC plot
plot(elc_obj, log_y = FALSE)
```

## 09.4.4 Expected value of perfect information (EVPI)

```{r}
evpi <- calc_evpi(wtp = v_wtp, psa = l_psa)
# EVPI plot
plot(evpi, effect_units = "QALY")
```
